Quantum Hoare Types

### 2019-10-31 paper/paper.tex
Add a todo package, start using stabilizer types
Add some partial examples

\usepackage[braket]{qcircuit}
\usepackage{mathpartir}
\usepackage{xspace}

@@

@@

@@

\begin{figure}[hb]
\begin{tabular}{lrcl}
	\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P $
	\end{minipage}\\
	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
	\end{minipage}\\
	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
		\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
	\end{minipage}\\
	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
\end{tabular}
	\caption{Syntax}
\label{fig:syntax}
\end{figure}

@@

\textit{Types.}

\subsubsection{Assertions}

\subsubsection{Example}

>>>>

\usepackage[braket]{qcircuit}
\usepackage{mathpartir}
\usepackage{xspace}
\usepackage[nowatermark]{fixmetodonotes}

@@

\newcommand{\Z}{\ensuremath{\mathbf{Z}}\xspace}
\newcommand{\X}{\ensuremath{\mathbf{X}}\xspace}
\newcommand{\E}{\ensuremath{\mathbf{E}}\xspace}
\renewcommand{\G}{\ensuremath{\mathbf{G}}\xspace}

\newcommand\hoare[3]{\{\,#1\,\}&#2&\{\,#3\,\}}

@@

\TODO{Figure out if stabilizer types fit well as assertions (need to frame them as containment relation, at the least)}

@@

\begin{figure}[hb]
\begin{tabular}{lrcl}
	\textit{Types} & $\tau, \sigma$ & ::= & $ \mathrm{qbit} \mid \tau \otimes \sigma \mid 1 \mid \Pi x{:}\tau.\sigma \mid \Psi. X. \{P\} x{:}\tau \{Q\}$ \\
	\textit{Assertions} & $P, Q, R$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{id}_\tau(M,N) \mid \mathrm{seleq}_\tau(H,M,N) \mid \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid \\
		P \supset Q \mid \neg P \mid \forall h{:}\mathrm{heap}.P \mid \Z \mid \X \mid \E \mid \G $
	\end{minipage}\\
	\textit{(Quantum) Heaps} & $H, G$ & ::= & $ h \mid \mathrm{empty} \mid \mathrm{upd}_{\tau}(H, M, N) $\\
	\textit{Elim terms} & $K, L$ & ::= & $ x \mid K\ M \mid M : \tau $\\
	\textit{Intro terms} & $M, N, O$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ K \mid (\,) \mid \lambda x.\ M \mid \mathrm{do}\ E \mid \ket{0} \mid \ket{1} \mid (M, N) \mid M==N $
	\end{minipage}\\
	\textit{Commands} & $c$ & ::= & \begin{minipage}[t]{0.6\columnwidth}%
		$ \mathrm{alloc}_\tau\ M \mid \mathrm{dealloc}\ M \mid \mathrm{if}_\tau\ M\ \mathrm{then}\ E_1\ \mathrm{else}\ E_2 \mid
		\mathrm{fix}\ f(y{:}\tau){:}\sigma = \mathrm{do}\ E\ \mathrm{in\ eval}\ f M \mid \mathrm{apply}(M, N) \mid \mathrm{capply}(M, N, O) $
	\end{minipage}\\
	\textit{Computations} & $E, F$ & ::= & $ \mathrm{return}\ M \mid x \gets K; E \mid x \Leftarrow c; E \mid x =_\tau M; E $\\
	\textit{Variable context} & $\Delta, \Psi$ & ::= & $ \cdot \mid \Delta, x{:}\tau $\\
	\textit{Heap context} & $X$ & ::= & $ \cdot \mid X, h $\\
	\textit{Assertion context} & $\Gamma$& ::= & $\cdot \mid \Gamma, P $
\end{tabular}
	\caption{Syntax}
\label{fig:syntax}
\end{figure}

@@

\textit{Types.} Primitive types such as qbit, pair, unit, dependent functions, and the Hoare type.

\textit{Assertions.} We introduce the notion of stabilizer types \Z, \X, \E, \G as type assertions. \TODO{shouldn't these be really treated as type refinements?}

\subsubsection{Example}
\begin{enumerate}
	\item Bell Pair:
\[ \begin{array}{rcl}
\hoare{a, b : \Z}{H(a);}{a : \X; b: \Z} \\
\hoare{a : \X; b: \Z}{CX(a, b);}{(a, b): \E} \\
\end{array} \]

	\item Superdense coding:
\[ \begin{array}{rcl}
\hoare{x, y : \Z; (a, b): \E}{CZ(x,a); \; CX(y, a); \; transmit(A, B);}{x, y : \Z; (a, b): \E} \\
\hoare{x, y : \Z; (a, b): \E}{CX(a, b);}{x, y : \Z; (a, b): \X \ast \Z} \\
\hoare{x, y, b : \Z; a: \X}{H(a);}{x, y, a, b : \Z} \\
\end{array} \]

or for the complete program:

\[ \begin{array}{rcl}
\hoare{x, y : \Z; (a, b): \E}{ superdense(x, y, a, b);}{x, y, a, b : \Z} \\
\end{array} \]

\TODO{figure out ownership types and pure vs mixed state types}

	\item Entanglement Swapping (variant of teleportation):\\
	Initially Alice ($a_1$) and Carol ($c$) share an EPR pair and so do Alice ($a_2$) and Bob ($b$)
\[ \begin{array}{rcl}
\hoare{(c, a_1), (a_2, b): \E}{CX(a_1, a_2);}{??} \\
\hoare{??}{H(a_1)}{??}\\
\hoare{??}{meas(a_1); \; meas(a_2);}{(c,b): \E; a_1, a_2 : qbit}\\
\end{array} \]

\item BB84 Protocol for key distribution:
\end{enumerate}
